---
title: "Modeling Practice"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
execute:
  freeze: auto
---

load libraries 

```{r}
library(tidyverse)
library(tidymodels)
library(lubridate)
library(glmnet)
library(rpart.plot)
library(vip)
library(baguette)
options(scipen=999)
```
read in the data 
```{r}
bike_data <- read_csv("https://www4.stat.ncsu.edu/~online/datasets/SeoulBikeData.csv",
local = locale(encoding = "latin1"))
bike_data
```

check the data for missing values

```{r}
bike_data |>
is.na()|>
colSums()
```

check column type and values

```{r}
attributes(bike_data)$spec
```

fix date column

```{r}
bike_data <- bike_data |>
 mutate(date = lubridate::dmy(Date)) |>
 select(-Date)
```

summarize each column

```{r}
summary(bike_data)
```

check character columns

```{r}
bike_data$Seasons |>
 unique()
```

```{r}
bike_data$Holiday |>
 unique()
```

```{r}
bike_data$`FunctioningDay`|>
 unique()
```

Turn them into factor variables

```{r}
bike_data<-bike_data |>
mutate(seasons=factor(Seasons),
holiday= factor(Holiday),
fn_day= factor(`Functioning Day`))|>
select(-Seasons,-Holiday,-`Functioning Day`)
```

rename variables

```{r}
 bike_data<-bike_data |>
 rename('bike_count' = `Rented Bike Count`,
 'hour' ="Hour",
 "temp"= `Temperature(°C)`,
 "wind_speed"= `Wind speed (m/s)`,
 "humidity"= `Humidity(%)`,
 "vis"= `Visibility (10m)`,
 "dew_point_temp"= `Dew point temperature(°C)`,
 "solar_radiation"= `Solar Radiation (MJ/m2)`,
 "rainfall"="Rainfall(mm)",
 "snowfall"= `Snowfall (cm)`)
```

remove observations from fn_day because they were out of commission those days

```{r}
 bike_data<-bike_data |>
 filter(fn_day =="Yes") |>
 select(-fn_day)
```

To simplify our analysis, we’ll summarize across the hours so that each day  has one observation associated with it. Let’s group_by() the date,seasons,and holiday variables and find the sumo f the bike_count, rainfall,and snowfall variables and the mean of all the weather related variables.

```{r}
 bike_data<-bike_data |>
 group_by(date,seasons,holiday)|>
 summarize(bike_count=sum(bike_count),
 temp= mean(temp),
 humidity= mean(humidity),
 wind_speed= mean(wind_speed),
 vis= mean(vis),
dew_point_temp= mean(dew_point_temp),
 solar_radiation= mean(solar_radiation),
 rainfall=sum(rainfall),
 snowfall=sum(snowfall))|>
 ungroup()
```

show cleaned data

```{r}
bike_data
```

**Summary Stats & Graphs**

```{r}
 bike_data |>
 summarize(across(`bike_count`,
 .fns= c("mean" = mean,
 "median"=median,
 "sd"=sd,
 "IQR"=IQR,
 "min"=min,
 "max"=max),
 .names="{.col}_{.fn}"))
```

```{r}
bike_data |>
 group_by(holiday)|>
 summarize(across(`bike_count`,
 .fns= c("mean" = mean,
 "median"=median,
 "sd"=sd,
 "IQR"=IQR,
 "min"=min,
 "max"=max),
 .names="{.col}_{.fn}"))
```

```{r}
bike_data |>
 group_by(seasons)|>
 summarize(across(`bike_count`,
 .fns= c("mean" = mean,
 "median"=median,
 "sd"=sd,
 "IQR"=IQR,
 "min"=min,
 "max"=max),
 .names="{.col}_{.fn}"))
```

correlations
```{r}
bike_data |>
 select(where(is.numeric))|>
 cor()|>
 round(3)
```

visualizations
```{r}
 ggplot(bike_data, aes(x=temp,y=bike_count))+
 geom_jitter(aes(color=seasons)) +
 facet_grid(~holiday)
```

```{r}
 ggplot(bike_data, aes(x = solar_radiation, y = bike_count)) +
 geom_point(aes(color = seasons)) +
 facet_grid(~holiday)
```

 Split the Data
 • Use functions from tidymodels to split the data into a training and test set (75/25 split). Use the
 strata argument to stratify the split on the seasons.
 • Onthe training set, create a 10 fold CV split
```{r}
 set.seed(11)
 bike_split <- initial_split(bike_data, prop = 0.75, strata = seasons)
 bike_train <- training(bike_split)
 bike_test <- testing(bike_split)
 bike_10_fold <- vfold_cv(bike_train, 10)
```
 
Fitting MLR Models

1st recipe 
```{r}
 MLR_rec1<-recipe(bike_count~ ., data=bike_train)|>
 step_date(date,features="dow")|>
 step_mutate(day_type=factor(if_else(date_dow %in%c("Sat","Sun"),"Weekend","Weekday"))) |>
 step_rm(date,date_dow) |>
 step_dummy(seasons,holiday,day_type) |>
 step_normalize(all_numeric(),-bike_count)
```

2nd recipe 
```{r}
 MLR_rec2<-MLR_rec1 |>
 step_interact(terms=~starts_with("seasons")*starts_with("holiday")+
 starts_with("seasons")*temp +
 temp*rainfall)
```

3rd recipe 
```{r}
 MLR_rec3<-MLR_rec2 |>
 step_poly(temp,
 wind_speed,
 vis,
 dew_point_temp,
 solar_radiation,
 rainfall,
 snowfall,
 degree=2)
```

set up linear model fit 
```{r}
MLR_spec<-linear_reg()|>
 set_engine("lm")
```

 Fitthemodelsusing10foldCVandconsiderthetrainingsetCVerrortochooseabestmodel.
```{r}
MLR_CV_fit1<-workflow()|>
 add_recipe(MLR_rec1) |>
 add_model(MLR_spec)|>
 fit_resamples(bike_10_fold)
 MLR_CV_fit2 <-workflow()|>
 add_recipe(MLR_rec2)|>
 add_model(MLR_spec)|>
 fit_resamples(bike_10_fold)
 MLR_CV_fit3 <-workflow()|>
 add_recipe(MLR_rec3) |>
 add_model(MLR_spec)|>
 fit_resamples(bike_10_fold)
```
 
get metrics
```{r}
 rbind(MLR_CV_fit1 |>collect_metrics(),
 MLR_CV_fit2 |>collect_metrics(),
 MLR_CV_fit3 |>collect_metrics())
```

last model seems to be the best. lets see how it performs on test data set
```{r}
 final_fit<-workflow()|>
 add_recipe(MLR_rec1) |>
 add_model(MLR_spec)|>
 last_fit(bike_split)
 final_fit |>
 collect_metrics()
```

 Obtain the final model (fit on the entire training set)coefficient table using tidy().
```{r}
final_fit |>
 extract_fit_parsnip()|>
 tidy()
```
 
 lasso regression 
```{r}
lasso_spec <- linear_reg(penalty = tune(), mixture = 1) |> set_engine("glmnet")

lasso_wf <- workflow() |>
add_recipe(MLR_rec1) |>
add_model(lasso_spec)

lasso_grid <- grid_regular(penalty(range = c(-4, 0)), levels = 20)

set.seed(123)
lasso_res <- tune_grid(
lasso_wf,
resamples = bike_10_fold,
grid = lasso_grid,
metrics = metric_set(rmse, mae)
)

best_lasso <- lasso_res |> select_best(metric = "rmse")

lasso_final <- finalize_workflow(lasso_wf, best_lasso) |>
last_fit(bike_split)

lasso_final |> collect_metrics()
lasso_final |> extract_fit_parsnip() |> tidy()

```
 
regression tree
```{r}
tree_spec <- decision_tree(cost_complexity = tune(), tree_depth = tune(), min_n = tune()) |>
set_engine("rpart") |>
set_mode("regression")

tree_wf <- workflow() |>
add_recipe(MLR_rec1) |>
add_model(tree_spec)

tree_grid <- grid_regular(
cost_complexity(range = c(0.001, 0.1)),
tree_depth(range = c(2, 10)),
min_n(range = c(5, 20)),
levels = 5
)

set.seed(123)
tree_res <- tune_grid(tree_wf, resamples = bike_10_fold, grid = tree_grid, metrics = metric_set(rmse, mae))
best_tree <- tree_res |> select_best(metric = "rmse")

tree_final <- finalize_workflow(tree_wf, best_tree) |>
last_fit(bike_split)

tree_final |> collect_metrics()

final_tree_fit <- tree_final |> extract_fit_parsnip() |> extract_fit_engine()
rpart.plot(final_tree_fit)

```

bagged tree 
```{r}
bag_spec <- bag_tree(
  min_n = tune(),
  tree_depth = tune()
) |>
  set_engine("rpart", times = 500) |>
  set_mode("regression")

bag_wf <- workflow() |>
  add_recipe(MLR_rec1) |>
  add_model(bag_spec)

bag_grid <- grid_regular(
  min_n(range = c(2, 20)),
  tree_depth(range = c(3, 10)),
  levels = 3
)

set.seed(123)
bag_res <- tune_grid(
  bag_wf,
  resamples = bike_10_fold,
  grid = bag_grid,
  metrics = metric_set(rmse, mae)
)

best_bag <- bag_res |> select_best(metric = "rmse")

bag_final <- finalize_workflow(bag_wf, best_bag) |>
  last_fit(bike_split)

bag_final |> collect_metrics()

bag_fit <- bag_final |>
  extract_fit_parsnip() |>
  extract_fit_engine()

vip(bag_fit)

```

Fit model to CV folds & view metrics across each fold

```{r}
bag_fit <- bag_wkf |>
  tune_grid(resamples = train_10_fold,
            grid = grid_regular(cost_complexity(), levels = 10),
            metrics = metric_set(rmse, mae))

bag_fit |>
  collect_metrics(type = 'wide') |>
  arrange(rmse, mae)
```

Select the model with the lowest RMSE to finalize

```{r}
bag_best <- bag_fit |>
  select_best(metric = 'rmse')
bag_best
```

Finalize the workflow using the best model and fit on the full training set

```{r}
bag_spec <- bag_tree(tree_depth = 10,
                     min_n = 6,
                     cost_complexity = tune()) |>
  set_engine('rpart') |>
  set_mode('regression')

# workflow
bag_wkf <- workflow() |>
  add_recipe(rec) |>
  add_model(bag_spec)
```

Fit model to CV folds & view metrics across each fold

```{r}
bag_fit <- bag_wkf |>
  tune_grid(resamples = train_10_fold,
            grid = grid_regular(cost_complexity(), levels = 10),
            metrics = metric_set(rmse, mae))

bag_fit |>
  collect_metrics(type = 'wide') |>
  arrange(rmse, mae)
```

Select the model with the lowest RMSE to finalize

```{r}
bag_best <- bag_fit |>
  select_best(metric = 'rmse')
bag_best
```

Finalize the workflow using the best model and fit on the full training set

```{r}
bag_final <- bag_wkf |>
  finalize_workflow(bag_best) |>
  last_fit(data_split, metrics = metric_set(rmse, mae))

bag_final |>
  collect_metrics()
```

random forest
```{r}
rf_spec <- rand_forest(trees = 500, min_n = tune(), mtry = tune()) |>
set_engine("ranger") |>
set_mode("regression")

rf_wf <- workflow() |>
add_recipe(MLR_rec1) |>
add_model(rf_spec)

rf_grid <- grid_regular(
min_n(range = c(2, 20)),
mtry(range = c(2, 8)),
levels = 3
)

set.seed(123)
rf_res <- tune_grid(rf_wf, resamples = bike_10_fold, grid = rf_grid, metrics = metric_set(rmse, mae))
best_rf   <- rf_res   |> select_best(metric = "rmse")

rf_final <- finalize_workflow(rf_wf, best_rf) |>
last_fit(bike_split)

rf_final |> collect_metrics()

rf_fit <- rf_final |> extract_fit_parsnip() |> extract_fit_engine()
vip(rf_fit)

```

compare all models 
```{r}
rbind(
  MLR_final |>
    collect_metrics(type = 'wide') |> 
    mutate(Model = 'MLR', .before = 'rmse'),
  LASSO_final |>
    collect_metrics(type = 'wide') |> 
    mutate(Model = 'LASSO', .before = 'rmse'),
  reg_tree_final |>
    collect_metrics(type = 'wide') |> 
    mutate(Model = 'TREE', .before = 'rmse'),
  bag_final |> 
    collect_metrics(type = 'wide') |> 
    mutate(Model = 'BAGGED', .before = 'rmse'),
  rf_final |>
    collect_metrics(type = 'wide') |> 
    mutate(Model = 'RF', .before = 'rmse')
) |>
  arrange(rmse, mae)
```

fit the overall best model on the entire dataset
```{r}
final_fit <- workflow() |>
  add_recipe(MLR_rec3) |>
  add_model(MLR_spec) |>
  last_fit(bike_split)
```

