---
title: "Modeling Practice"
format: html
editor: visual
---

read in the data

```{r}
library(tidyverse)
library(tidymodels)
bike_data <- read_csv("https://www4.stat.ncsu.edu/~online/datasets/SeoulBikeData.csv",
local = locale(encoding = "latin1"))
bike_data
```

check the data for missing values

```{r}
bike_data |>
is.na()|>
colSums()
```

check column type and values

```{r}
attributes(bike_data)$spec
```

fix date column

```{r}
bike_data <- bike_data |>
 mutate(date = lubridate::dmy(Date)) |>
 select(-Date)
```

summarize each column

```{r}
summary(bike_data)
```

check character columns

```{r}
bike_data$Seasons |>
 unique()
```

```{r}
bike_data$Holiday |>
 unique()
```

```{r}
bike_data$`FunctioningDay`|>
 unique()
```

Turn them into factor variables

```{r}
 bike_data<-bike_data |>
 mutate(seasons=factor(Seasons),
 holiday= factor(Holiday),
 fn_day= factor(`FunctioningDay`))|>
 select(-Seasons,-Holiday,-`FunctioningDay`)
```

rename variables

```{r}
 bike_data<-bike_data |>
 rename('bike_count' = `RentedBike Count`,
 'hour' ="Hour",
 "temp"= `Temperature(°C)`,
 "wind_speed"= `Windspeed(m/s)`,
 "humidity"= `Humidity(%)`,
 "vis"= `Visibility(10m)`,
 "dew_point_temp"= `Dewpointtemperature(°C)`,
 "solar_radiation"= `SolarRadiation(MJ/m2)`,
 "rainfall"="Rainfall(mm)",
 "snowfall"= `Snowfall(cm)`)
```

remove observations from fn_day because they were out of commission those days

```{r}
 bike_data<-bike_data |>
 filter(fn_day =="Yes") |>
 select(-fn_day)
```

Tosimplifyouranalysis,we’ll summarizeacrossthehourssothateachdayhasoneobservationassociated withit. Let’sgroup_by()thedate,seasons,andholidayvariablesandfindthesumof thebike_count, rainfall,andsnowfallvariablesandthemeanofall theweatherrelatedvariables.

```{r}
 bike_data<-bike_data |>
 group_by(date,seasons,holiday)|>
 summarize(bike_count=sum(bike_count),
 temp= mean(temp),
 humidity= mean(humidity),
 wind_speed= mean(wind_speed),
 vis= mean(vis),
dew_point_temp= mean(dew_point_temp),
 solar_radiation= mean(solar_radiation),
 rainfall=sum(rainfall),
 snowfall=sum(snowfall))|>
 ungroup()
```

show cleaned data

```{r}
bike_data
```

**Summary Stats & Graphs**

```{r}
 bike_data |>
 summarize(across(`bike_count`,
 .fns= c("mean" = mean,
 "median"=median,
 "sd"=sd,
 "IQR"=IQR,
 "min"=min,
 "max"=max),
 .names="{.col}_{.fn}"))
```

```{r}
bike_data |>
 group_by(holiday)|>
 summarize(across(`bike_count`,
 .fns= c("mean" = mean,
 "median"=median,
 "sd"=sd,
 "IQR"=IQR,
 "min"=min,
 "max"=max),
 .names="{.col}_{.fn}"))
```

```{r}
bike_data |>
 group_by(seasons)|>
 summarize(across(`bike_count`,
 .fns= c("mean" = mean,
 "median"=median,
 "sd"=sd,
 "IQR"=IQR,
 "min"=min,
 "max"=max),
 .names="{.col}_{.fn}"))
```

correlations
```{r}
bike_data |>
 select(where(is.numeric))|>
 cor()|>
 round(3)
```

visualizations
```{r}
 ggplot(bike_data, aes(x=temp,y=bike_count))+
 geom_jitter(aes(color=seasons)) +
 facet_grid(~holiday)
```

```{r}
 ggplot(bike_data, aes(x = solar_radiation, y = bike_count)) +
 geom_point(aes(color = seasons)) +
 facet_grid(~holiday)
```

 Split the Data
 • Use functions from tidymodels to split the data into a training and test set (75/25 split). Use the
 strata argument to stratify the split on the seasons.
 • Onthe training set, create a 10 fold CV split
```{r}
 set.seed(11)
 bike_split <- initial_split(bike_data, prop = 0.75, strata = seasons)
 bike_train <- training(bike_split)
 bike_test <- testing(bike_split)
 bike_10_fold <- vfold_cv(bike_train, 10)
```
 
Fitting MLR Models

1st recipe 
```{r}
 MLR_rec1<-recipe(bike_count~ ., data=bike_train)|>
 step_date(date,features="dow")|>
 step_mutate(day_type=factor(if_else(date_dow %in%c("Sat","Sun"),"Weekend","Weekday"))) |>
 step_rm(date,date_dow) |>
 step_dummy(seasons,holiday,day_type) |>
 step_normalize(all_numeric(),-bike_count)
```

2nd recipe 
```{r}
 MLR_rec2<-MLR_rec1 |>
 step_interact(terms=~starts_with("seasons")*starts_with("holiday")+
 starts_with("seasons")*temp +
 temp*rainfall)
```

3rd recipe 
```{r}
 MLR_rec3<-MLR_rec2 |>
 step_poly(temp,
 wind_speed,
 vis,
 dew_point_temp,
 solar_radiation,
 rainfall,
 snowfall,
 degree=2)
```

set up linear model fit 
```{r}
MLR_spec<-linear_reg()|>
 set_engine("lm")
```

 Fitthemodelsusing10foldCVandconsiderthetrainingsetCVerrortochooseabestmodel.
```{r}
MLR_CV_fit1<-workflow()|>
 add_recipe(MLR_rec1) |>
 add_model(MLR_spec)|>
 fit_resamples(bike_10_fold)
 MLR_CV_fit2 <-workflow()|>
 add_recipe(MLR_rec2)|>
 add_model(MLR_spec)|>
 fit_resamples(bike_10_fold)
 MLR_CV_fit3 <-workflow()|>
 add_recipe(MLR_rec3) |>
 add_model(MLR_spec)|>
 fit_resamples(bike_10_fold)
```
 
get metrics
```{r}
 rbind(MLR_CV_fit1 |>collect_metrics(),
 MLR_CV_fit2 |>collect_metrics(),
 MLR_CV_fit3 |>collect_metrics())
```

last model seems to be the best. lets see how it performs on test data set
```{r}
 final_fit<-workflow()|>
 add_recipe(MLR_rec1) |>
 add_model(MLR_spec)|>
 last_fit(bike_split)
 final_fit |>
 collect_metrics()
```

 Obtainthefinalmodel (fitontheentiretrainingset)coefficienttableusingtidy().
```{r}
final_fit |>
 extract_fit_parsnip()|>
 tidy()
```
 
 